### 2021-04-07 pldi21/livelit-definitions.tex
second round

The expansion can include the spliced expressions, 
but the system does not make the expression  
available directly as a value of type \li{Exp}. 
Instead, \li{expand} returns a \emph{proto-expansion}, of type \li{Exp}, 
paired with a list of \li{SpliceRef}s (which can come from 
both splices and parameters as discussed above).
The {proto-expansion}
 is an encoding (i.e. a value of type \li{Exp}) of a function 
 (here curried)
 that takes an argument for each listed \li{SpliceRef}. 
 That argument is of the corresponding splice 
type, which was provided when the splice was initialized. 
 The return type of the proto-expansion is the expansion type.
 So here, the proto-expansion for \li{\$color} must be a function of type 
 \li{Int -> Int -> Int -> Int -> Color}.

Note that Hazel does \emph{not} statically check the definition of \li{expand}
to ensure that the encoded proto-expansion has this type. Instead, the proto-expansion 
is \emph{validated} at each livelit invocation site, with errors reported to the client. 
A \emph{typed quotation} system as in, e.g., MetaOCaml \cite{DBLP:conf/flops/Kiselyov14}, could be adapted to allow for definition-site verification 
in lieu of client-site validation, as discussed in \cite{TLMs}, but we leave the integration 
of such a system into Hazel as future work. This would be non-trivial:  
the type of the quotation depends on the type of each splice in 
the splice list, which may be of model-dependent length (e.g. in the case of \li{\$dataframe}).

This parameterization strategy makes it simple to enforce the binding discipline described in Sec.~\ref{sec:hygiene}. 
The proto-expansion 
can depend only on the \li{context}, whereas the splices are entered by the client and so they can 
depend only on the client site typing context. The use of function application ensures
that splices are capture avoiding as discussed in Sec.~\ref{sec:hygiene}. 
We consider this more formally next.

>>>>

The expansion can include the spliced expressions, 
but the system does not make the expression  
available directly as a value of type \li{Exp}. 
Instead, \li{expand} returns a \emph{parameterized expansion}, of type \li{Exp}, 
paired with a list of \li{SpliceRef}s (which can come from 
both splices and parameters as discussed above).
The {parameterized expansion}
 is an encoding (i.e. a value of type \li{Exp}) of a function 
 (here curried)
 that takes an argument for each listed \li{SpliceRef}. 
 That argument is of the corresponding splice 
type, which was provided when the splice was initialized. 
 The return type of the parameterized expansion is the expansion type.
 So here, the parameterized expansion for \li{\$color} must be a function of type 
 \li{Int -> Int -> Int -> Int -> Color}.

Note that Hazel does \emph{not} statically check the definition of \li{expand}
to ensure that the encoded parameterized expansion has this type. Instead, the parameterized expansion 
is \emph{validated} at each livelit invocation site, with errors reported to the client. 
A \emph{typed quotation} system as in, e.g., MetaOCaml \cite{DBLP:conf/flops/Kiselyov14}, could be adapted to allow for definition-site verification, as discussed in \cite{TLMs}, but 
note that   
the type of the quotation depends on the type of each splice in 
the splice list.

This parameterization strategy makes it simple to enforce the binding discipline described in Sec.~\ref{sec:hygiene}. 
The parameterized expansion 
can depend only on the \li{context}, whereas the splices are entered by the client and so they can 
depend only on the client site typing context. The use of function application ensures
that splices are capture avoiding as discussed in Sec.~\ref{sec:hygiene}. 
We consider this more formally next.

### 2021-04-07 pldi21/related-work.tex
second round

Recent work on a interactive visual macro system in Racket 
is quite similar in spirit to our work \cite{interactive-visual-syntax}.
However, it supports only a limited form of splicing via a pop up text editor
that does not support full compositionality as described here, where livelits 
can appear within other livelits. Splicing is not strictly hygienic, though 
Racket's scope management machine can be used by macros to voluntarily 
maintain a binding discipline \cite{DBLP:conf/popl/Flatt16}. There is not any 
consideration of typing. Parameterization and partial application is also impossible, because invocation 
is via an editor command rather than a syntactic mechanism. Finally,
there is no support for liveness: macro evaluation occurs in the editor 
environment, not the run-time environment of the program being written, 
so splice evaluation would not work correctly (except in trivial cases
where the splice was, e.g., a closed expression, or used only standard 
library constructs).

@@

There has been a long line of research on \emph{projectional editing}, where
the user edits graphical representations (projections) of code constructs 
\cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. Livelits are a form of projectional editing. 
Many of the oldest systems offer only a fixed set of projections and interaction techniques. 
More recently, language workbenches with support for projectional editing like Citrus \cite{DBLP:conf/uist/KoM05}
and MPS \cite{voelter2011language} have made it easier to define new projectional editors.
However, these systems generate entire editors, whereas livelit definitions 
are defined in libraries. 
Furthermore, livelit definitions are governed by a type and binding discipline.
% In many ways, an unfamiliar livelit can be approached much like an unfamiliar function
% can be approached: by inspecting its declarations and reasoning compositionally 
% about arguments, i.e. splices and parameters.
Finally, livelits are uniquely live, building on Hazelnut Live \cite{HazelnutLive}.

@@

Related to projectional editors are 
a variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey}. Livelits differ 
from these systems in directionality: visualization systems are given code and generate
a visualization, whereas livelits are generate code where there would otherwise
be a hole.

>>>>

Recent work on a interactive visual macro system in Dr. Racket 
is quite similar in spirit to our work \cite{interactive-visual-syntax}.
However, it supports only a limited form of splicing via a pop up text editor
that does not support full compositionality as described here, where livelits 
can appear within other livelits. Splicing is not strictly hygienic, though 
Racket's scope management machinery can be used by macros to voluntarily 
maintain a binding discipline \cite{DBLP:conf/popl/Flatt16}. There is not any 
consideration of typing. Parameterization and partial application is also impossible, because invocation 
is via an editor command rather than a syntactic mechanism. Finally,
there is no support for liveness: macro evaluation occurs in the editor 
environment, not the run-time environment of the program being written, 
so splice evaluation would not work correctly (except in trivial cases
where the splice was, e.g., a closed expression, or used only standard 
library constructs). That said, we conjecture that it is
likely possible to implement many of the mechanisms we describe in this paper 
perhaps in some form as a layer atop existing mechanisms in Dr. Racket.

@@

There has been a long line of research on \emph{projectional editing}, where
the user edits graphical representations (projections) of code constructs 
\cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. Livelits are a form of projectional editing. 
Many of the oldest systems offer only a fixed set of projections and interaction techniques. 
More recently, language workbenches with support for projectional editing like Citrus \cite{DBLP:conf/uist/KoM05}
and MPS \cite{voelter2011language} have made it easier to define new projectional editors.
However, these systems generate entire editors, whereas livelit definitions 
are decentralized in libraries. 
Furthermore, livelit definitions are governed by a type and binding discipline.
% In many ways, an unfamiliar livelit can be approached much like an unfamiliar function
% can be approached: by inspecting its declarations and reasoning compositionally 
% about arguments, i.e. splices and parameters.
Finally, livelits are uniquely live, building on Hazelnut Live \cite{HazelnutLive}.

@@

Related to projectional editors are 
a variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey}. Livelits differ 
from these systems in directionality: visualization systems are given code and generate
a visualization, whereas livelits generate code where there would otherwise
be a hole.
